---
title: "Data Journalism Lesson 16: Waffle charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: journal
    ace_theme: github
runtime: shiny_prerendered
description: >
  Making quick graphics for reporting.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(glue)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

```{r load-data, message=FALSE, warning=FALSE}
rows <- 10

house <- data.frame(
  category = c("Republicans", "Democrats", "Vacant"),
  count = c(218, 213, 4)
)

waffle_house <- house |>
  uncount(count) |>
  mutate(
    x = (row_number() - 1) %/% rows + 1,
    y = (row_number() - 1) %% rows + 1
  )

state <- Sys.getenv("tutorial.state")
if(state == "") state <- "NE"

stateName <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/states.csv") |> filter(Postal == state) 

legislatures <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/state-legislatures/legislatures.csv")

xgr <- legislatures |> filter(state == stateName$State)

xgrrows <- nrow(xgr)

senate <- xgr |> 
  select(state, senate_dem, senate_rep, senate_other, senate_vacant) |>
  pivot_longer(
    cols=c(-state),
    names_to = "party",
    values_to = "members"
  ) |> 
  uncount(members) |>
  mutate(
    x = (row_number() - 1) %/% rows + 1,
    y = (row_number() - 1) %% rows + 1
  ) |> 
  ungroup()

mostpartisan <- legislatures |> 
  mutate(partisan_lean = 
           case_when(
             senate_dem > senate_rep ~ senate_dem/total_senate,
             senate_dem < senate_rep ~ senate_rep/total_senate
             )
         ) |> 
  top_n(10, wt=partisan_lean) |> 
  select(state, starts_with("senate")) |> 
  pivot_longer(
    cols=-state,
    names_to = "party",
    values_to = "members"
  ) |> 
  uncount(members) |>
  group_by(state) |> 
  mutate(
    x = (row_number() - 1) %/% rows + 1,
    y = (row_number() - 1) %% rows + 1
  ) |> 
  ungroup()

if (stateName$State %in% mostpartisan$state) {
  xgrs <- mostpartisan
} else {
  xgrs <- bind_rows(mostpartisan, senate)
}

```

## The Goal

In this lesson, you'll learn how to create waffle charts, an effective alternative to pie charts for visualizing proportions. By the end of this tutorial, you'll understand how to use ggplot to construct single and comparative waffle charts. You'll practice creating these charts with real data from legislatures, learning how to adjust chart parameters for clarity. This skill will enable you to present proportional data in a clear, visually appealing format that's easier to interpret than traditional pie charts.

## Why Visualize Data?

Show me someone who has a job in data visualization, and I'll show you someone who has Edward Tufte's *The Visualization of Quantitative Information* on their bookshelf. It's one of the most popular, and most influential books on data visualization ever written. As such, it's one of the most referenced, and most criticized, books on the subject. If visualizing data is interesting to you, you have to read it.

What will you find? A clearly articulated argument of what makes a data graphic good. So clearly articulated that some call it stifling, rigid, inflexible. Some of his commandments can be taken to extremes, where instead of intelligent and clear graphical excellence, as he describes it, you get something so esoteric it doesn't make sense.

But once you realize Tufte -- in spite of his unambiguous language -- is more like the pirate's code in Pirate's of the Caribbean (more of a guideline than a rule), then you'll see the genius of it all.

Indeed, page 1, part 1, gives you a taste of his direct style.

"EXCELLENCE in statistical graphics consists of complex ideas communicated with clarity, precision, and efficiency. Graphical displays should

-   show the data
-   induce the viewer to think about the substance rather than about methodology, graphic design, the technology of graphic production, or something else
-   avoid distorting what the data have to say
-   present many numbers in a small space
-   make large data sets coherent
-   encourage the eye to compare different pieces of data
-   reveal the data at several levels of detail, from a broad overview to the fine structure
-   serve a reasonably clear purpose: description, exploration, tabulation, or decoration
-   be closely integrated with the statistical and verbal descriptions of a data set. "

If you can do all that, you're on a great path. Easy right? Not exactly. Doing some of this is easy with the tools we have. Some of this requires a lot of thought.

That brings us to today when we come not to praise the pie chart but to bury it. They're terrible, and people abuse them horribly. Using Tufte's own words, pie charts regularly fail to show the data, often distorting it, and are truly horrible at making large datasets coherent.

This can lead to what Tufte calls failures of Graphical Integrity. Tufte insists that the early years of graphics were plagued with designers who used them to lie -- intentionally or not -- by covering up bad data or bad results with charts. He credits no less than John Tukey for making graphics respectable again in the late 1960s.

I'm here to argue that pie charts are the tool of the liar. But showing proportion of a whole is a very important job for a graphic. Fortunately, there are better tools. That tool is called the waffle chart, and we introduce those here.

As you're working today, keep Tufte's six principles of graphical integrity in mind:

"Graphical integrity is more likely to result if these six principles are followed:

-   The representation of numbers, as physically measured on the surface of the graphic itself, should be directly proportional to the numerical quantities represented.
-   Clear, detailed, and thorough labeling should be used to defeat graphical distortion and ambiguity. Write out explanations of the data on the graphic itself. Label important events in the data.
-   Show data variation, not design variation.
-   In time-series displays of money, deflated and standardized units of monetary measurement are nearly always better than nominal units.
-   The number of information-carrying (variable) dimensions depicted should not exceed the number of dimensions in the
-   Graphics must not quote data out of context."

## The Basics

Pie charts are the devil. They should be an instant F in any data visualization class. The problem with them? How carefully can you evaluate angles and area? Unless they are blindingly obvious and only a few categories, not well. If you've got 25 categories, how can you tell the difference between 7 and 9 percent? You can't.

So let's introduce a better way: The waffle chart. Why a waffle chart? Because it's made up of small squares, just like a waffle. Some call it a square pie chart. I personally hate that. Waffles it is.

A waffle chart is similar to a stacked bar chart. **A waffle chart is designed to show you parts of the whole -- proportionality**. How many things make up the whole. It will also show magnitude, just like a stacked bar.

What's the difference?

The truth is, there really isn't one. If I had to stand before a judge and come up with an answer or go to jail, I'd say that a stacked bar chart can get more things in it -- more counties, more states, more schools, etc. The waffle chart demands focus -- which you'll see in a minute -- and so you can really only handle a small number of waffle charts at once.

What that boils down to is that a waffle chart is a visually different form that accomplishes the same goal as a stacked bar chart, but works best when there's only a couple of things to compare. A two-bar stacked bar chart would look silly. A two-stack waffle chart would look great.

If you Google making waffle charts in R, you're going to find a library called `waffle` that does this pretty easily. Worse, if you ask AI to do it for you -- don't, that's an academic integrity violation -- it will use the waffle library do this. Problem? The library is abandoned. It won't install (easily) and worse, will slowly break in not-easily-understood ways. 

This library breaking is a new thing, so you're among the first to do it with just ggplot. And it's not that tough to pull off. We just need the `tidyverse`. 

To start, we load the `tidyverse`.

```{r load-waffle, exercise=TRUE}
library(tidyverse)
```

```{r load-waffle-solution}
library(tidyverse)
```

```{r load-waffle-check}
grade_this_code()
```

We're going to try this waffle chart several different ways. 

The easiest way to do waffle charts is to create your data manually. 

For this, we're going to look at the partisan breakdown of the U.S. House of Representatives. The House, without editorializing, is a mess, precisely because it's almost evenly divided (history nerd note: The House is supposed to be a mess. It was designed that way by the Founders. This is nothing new). For any measure to pass the House, the majority party has to have all of its members on board. They can't have people voting against. So what does that look visually?

First, we'll make our dataframe. The pattern is making a category and then inputting the data. Once you see it, it makes more sense. So, as of this writing, there's 218 Republican members of the House, 213 Democrats and 4 seats are vacant.

```{r load-data-exercise, exercise = TRUE}
house <- data.frame(
  category = c("Republicans", "Democrats", "Vacant"),
  count = c(218, 213, 4)
)
```

```{r load-data-exercise-solution}
house <- data.frame(
  category = c("Republicans", "Democrats", "Vacant"),
  count = c(218, 213, 4)
)
```

```{r load-data-exercise-check}
grade_this_code()
```

Next, we have to turn our data into a grid. 

### Exercise 1: Make a grid

Now we're going to create a dataframe called `waffle_house` -- get it? It's a waffle chart about the House of Representatives? Thus `waffle_house`. What is going into it and how it gets made is like this, in this order: 

* First, we set the number of rows. Because we have a lot of squares, we're going to set the number of rows at 10. 
* Then, start our data analysis. `waffle_house` is going to pull our data from `house`.
* Next we are going to use the uncount function, which is the opposite of the count function. If `house` had a party with 10 members, then uncount turns that into 10 rows of that party. 
* Next we have to create rows for our waffle chart. We're creating squares that will go left and right on the x-axis and up and down on the y-axis in a grid. 
* `row_number()` is just what you think it is -- what row number is it? Remember uncount made every member into a row. 
* For the x axis in mutate, `%/%` means divide and drop the remainder. In practice, what that says is if there's 30 members, the first 10 will be in row 1, the next 10 will be in row 2 and so forth as specified by the number of rows set above.
* For the y axis, `%%` means do the division and keep the remainder. In practice, that's going to increment the column up one each time. Row 1 column 1. Row 1 column 2. Row 1 column 3 and so on. 
* The + 1 parts just increments everything so the counts start at 1 not 0. 

```{r waffle, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
rows <- ____

waffle_house <- house |>
  uncount(count) |>
  mutate(
    x = (row_number() - 1) %/% rows + 1,
    y = (row_number() - 1) %% rows + 1
  )
```

```{r waffle-solution, exercise.reveal_solution = FALSE}
rows <- 10

waffle_house <- house |>
  uncount(count) |>
  mutate(
    x = (row_number() - 1) %/% rows + 1,
    y = (row_number() - 1) %% rows + 1
  )
```

```{r waffle-check}
grade_this_code()
```

Now we have a grid made up of category, an x coordinate and a y coordinate. 

### Exercise 2: Your first waffle 

We make waffles in ggplot with `geom_tile`, which does what you think it does -- it makes tiles, like a floor. `geom_tile` takes an `x`, a `y` and a fill to give it color. Good news -- we have an x, y and category column, which contains the party. 

What comes after is what makes this into a waffle chart. The `color` in this case creates a white line around the tiles and the `linewidth` does what you think it does. `coord_equal` makes all the squares an equal size. Then `theme_void` removes all of the x and y tick marks and labeling so all you see are squares.  We'll add a `scale_fill_manual` to make the colors what we want them to be. 

```{r waffle2, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() +
  geom_tile(
    data=waffle_house, 
    aes(
      x = ____, y = ____, fill = ____
      ), 
    color = "white", 
    linewidth = .5) +
  coord_equal() + 
  theme_void() + 
  scale_fill_manual(values = c("blue", "red", "grey"))
```

```{r waffle2-solution, exercise.reveal_solution = FALSE}
ggplot() +
  geom_tile(
    data=waffle_house, 
    aes(
      x = x, y = y, fill = category
      ), 
    color = "white", 
    linewidth = .5) +
  coord_equal() + 
  theme_void() + 
  scale_fill_manual(values = c("blue", "red", "grey"))
```

```{r waffle2-check}
grade_this_code()
```

Boom. You've made a waffle.

## Make more waffles 

What does it look like if we try to use a dataframe of data instead of one we create? It requires a little more work, but not a lot.

First, for this to work, your data needs to be long data, and you want as little of it as you can get.

```{r storysetup, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
if (stateName$State == "Nebraska") {
  glue("Instead of Congress, let's look at the partisan breakdown of the {stateName$State} Legislature. Now, technically speaking, the Nebraska Legislature is weird. First, the data is going to call it the state senate, but Nebraska only has one chamber (and its members are called senators). Even more weird: Nebraska is officially non-partisan. But which party each member belongs to is an open secret, so I've added the partisan breakdown to the National Conference of State Legislature's data on partisan control of state legislatures.")
} else {
  "Instead of Congress, let's look at the partisan breakdown of the {stateName$State} Senate. Why the senate and not the house? Well, not every state has a house (cough, cough Nebraska), so to make it easier, we'll focus on the chamber everyone has. This data comes from the National Conference of State Legislature's data on partisan control of state legislatures."
}
```

We're going to grab the data and filter all in one step. You just need to add your state name to the filter.

```{r xgr-read, exercise=TRUE, exercise.reveal_solution = FALSE, exercise.setup = "load-data"}
legislatures <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/state-legislatures/legislatures.csv")

xgr <- legislatures |> filter(state == "____")
```

```{r xgr-read-check}
grade_this({
  if (identical(nrow(.result), xgrrows)) {
    pass("Great work! You've got your state's legislature.")
  }
  fail()
})
```

Let's take a quick look at it. It won't take long -- it's one row of data. FYI: The data name `xgr` is one of those old-timey journalism terms from way back in the day. In the Times of Yore, legislature stories from the AP all started with the slug XGR.

```{r head-data, exercise=TRUE, exercise.setup = "load-data"}
head(____)
```

```{r head-data-solution, exercise.reveal_solution = FALSE}
head(xgr)
```

```{r head-data-check}
grade_this_code()
```

The problem we have is that it's one row -- and it's got lots of things we don't want or need. First, we want to get rid of all columns that aren't senate related. We can do that with select and just add each column we want. Then, we'll use `pivot_longer` -- the twin sibling of `pivot_wider` -- to make our wide data long. With `pivot_longer`, you have to tell it what columns are getting pivoted. You can just name them. But in our case, since we're pivoting every column *except* state, we can use a trick. If you put a negative sign in front of a column in cols, what you're saying is pivot everything EXCEPT this one. Then, instead of `names_from` and `values_from` like we used with `pivot_wider`, here we are giving names to two new columns. And because we're giving the names with `names_to` and `values_to` we can make these whatever we want. The first one is going to be the party, the second is the number of members in that party. Here's what it looks like.

### Exercise 3: Another grid 

After that, we can just add the same uncount and mutate as before, just adjusting uncount to be our members column. 

```{r waffle5, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
senate <- xgr |> 
  select(state, senate_dem, senate_rep, senate_other, senate_vacant) |>
  pivot_longer(
    cols=c(-state),
    names_to = "party",
    values_to = "members"
  ) |> 
  uncount(____) |>
  mutate(
    x = (row_number() - 1) %/% rows + 1,
    y = (row_number() - 1) %% rows + 1
  ) |> 
  ungroup()
```

```{r waffle5-solution, exercise.reveal_solution = FALSE}
senate <- xgr |> 
  select(state, senate_dem, senate_rep, senate_other, senate_vacant) |>
  pivot_longer(
    cols=c(-state),
    names_to = "party",
    values_to = "members"
  ) |> 
  uncount(members) |>
  mutate(
    x = (row_number() - 1) %/% rows + 1,
    y = (row_number() - 1) %% rows + 1
  ) |> 
  ungroup()
```

```{r waffle5-check}
grade_this_code()
```

Now take a look at `senate`:

```{r head2-data, exercise=TRUE, exercise.setup = "load-data"}
head(senate)
```

```{r head2-data-solution, exercise.reveal_solution = FALSE}
head(senate)
```

```{r head2-data-check}
grade_this_code()
```

Now we can make another waffle. 

### Exercise 4: Your legislature

What changes here? Not much. We want to show the breakdown by party, so that means we have a different fill from before and the order of the colors in `scale_fill_manual` changes because the alphabetical order is different. Otherwise, it's the same. 

```{r waffle6, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() +
  geom_tile(
    data=senate, 
    aes(
      x = x, y = y, fill = ____
      ), 
    color = "white", 
    linewidth = .5) +
  coord_equal() + 
  theme_void() + 
  scale_fill_manual(values = c("blue", "grey", "red"))
```

```{r waffle6-solution, exercise.reveal_solution = FALSE}
ggplot() +
  geom_tile(
    data=senate, 
    aes(
      x = x, y = y, fill = party
      ), 
    color = "white", 
    linewidth = .5) +
  coord_equal() + 
  theme_void() + 
  scale_fill_manual(values = c("blue", "grey", "red"))
```

```{r waffle6-check}
grade_this_code()
```

Now we have a similar -- but not quite exact -- copy of what we did with the House, except now for your state's senate. Which party has the majority? Is that clear from your chart?

### Exercise 5: Lots of waffles, one chart

What if we want to look at multiple states? We need to add one thing to our grid data from your state: A group by. If we group by the state and then do the mutates, then each state gets their own grid coordinates. 

```{r storysetup2, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("Behind the scenes, I've created a dataframe called `xgrs` that has the most lopsided partisan legislatures in the nation so we can compare {stateName$State} to those. My definition of lopsided partisan -- take the party with the majority, divide it by the total number of seats. Basically, the percentage of seats to that party. Arrange by that percentage and take the top 10. What you get are the 10 bluest or reddest state senates in the country.")
```

So how does yours compare?

The trick here is using something called faceting. Faceting makes lots of little charts based on a category -- like a state, or year, or something like that. With `geom_tile` before, we made one waffle with one state. With faceting, we can make a waffle for *each* state, faceting on the state name.

Let's take a quick peek at `xgrs` before we try to chart this.

```{r head3-data, exercise=TRUE, exercise.setup = "load-data"}
head(____)
```

```{r head3-data-solution, exercise.reveal_solution = FALSE}
head(xgrs)
```

```{r head3-data-check}
grade_this_code()
```

Just like our previous dataframe, there's a `party` column. What we need to `facet_wrap` -- that's the `state`. 

```{r waffle7, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() +
  geom_tile(
    data=mostpartisan, 
    aes(
      x = x, y = y, fill = ____
      ), 
    color = "white", 
    linewidth = .5) +
  facet_wrap(~____, ncol = 5, strip.position = "bottom") +
  coord_equal() + 
  theme_void() + 
  scale_fill_manual(values = c("blue", "red", "grey"))
```

```{r waffle7-solution, exercise.reveal_solution = FALSE}
ggplot() +
  geom_tile(
    data=mostpartisan, 
    aes(
      x = x, y = y, fill = party
      ), 
    color = "white", 
    linewidth = .5) +
  facet_wrap(~state, ncol = 5, strip.position = "bottom") +
  coord_equal() + 
  theme_void() + 
  scale_fill_manual(values = c("blue", "red", "grey"))
```

```{r waffle7-check}
grade_this_code()
```

Note there's a few extra bits in `facet_wrap` that helps us here. First -- `ncol`. What's strange about ncol in this case is that it puts the waffles 5 to a *row*. I don't get it either. The `strip.position` part just says where to put the label for each facet -- that's the state name.

What do we see here? Well, we have some work to do to bring this to publication. Headlines and explanation text primarily. But it's doing pretty well.

There are much bigger questions you should be thinking about now. Critical questions abound, questions you are going to have to confront regularly when making graphics:

1.  What story does this tell?
2.  Is that story clear?
3.  Is this the best way to tell this story?

If the answer is no -- to any of these, but especially 2 -- then you're going to have to start thinking about changing course. Does this chart tell a story? It's pretty clear there are some lop-sided senates in the United States. Is your state one of them?

**Just because you wrote code to make a graphic doesn't mean you have a graphic.** In journalism, the story is always the most important part. If it's not clear, if a reader who knows nothing about partisanship in America can't look at this and know what you're trying to communicate immediately, you need to keep working. That might be more fussing with this one. That might be dumping it entirely and trying something else.

## The Recap

Throughout this lesson, you've learned how to create and customize waffle charts using R's waffle library. You've practiced making single waffle charts to display proportions within a dataset, such as the partisan makeup of the U.S. House of Representatives. You've also learned how to create comparative waffle charts using `geom_waffle`, allowing you to visually compare proportions across different facets. You've discovered how to adjust your data and chart parameters to handle issues like scale differences and visual clarity. Remember, waffle charts offer a more precise and visually appealing alternative to pie charts for displaying proportional data, making them a valuable tool in your data visualization toolkit.

## Terms to Know

-   Waffle chart: A chart type that represents data as squares in a grid, where each square represents a proportion of the whole.
-   Vector: In R, a basic data structure that holds elements of the same type in a specific order.
-   `waffle()` function: The main function in the `waffle` library used to create a single waffle chart.
-   `geom_waffle()` function: A `ggplot` geom that makes waffle charts, either with a named vector or with a dataframe.
-   Concatenate (`c()`): An R function used to create vectors by combining elements.
-   `rows` parameter: In waffle charts, determines the number of rows in the grid of squares.
-   `xlab`: A parameter in the waffle() function used to label the x-axis, often explaining what each square represents.
-   `colors` parameter: Used in waffle charts to specify the colors for each category in the data.
-   Division in waffle charts: A technique used to reduce the number of squares by making each square represent multiple units of data.
