---
title: "Data Journalism Lesson 19: Dumbbell charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: journal
    ace_theme: github
runtime: shiny_prerendered
description: >
  Making quick graphics for publication
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(glue)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```
```{r load-data, message=FALSE, warning=FALSE}
state <- Sys.getenv("tutorial.state")
if(state == "") state <- "NE"

stateName <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/states.csv") |> filter(Postal == state) 

stateName <- stateName |> 
  mutate(dataurl = case_when(
    Postal == "AL" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "AK" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "AZ" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "AR" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "CA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "CO" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "CT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "DE" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "FL" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "GA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "HI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "ID" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "IL" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "IN" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "IA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "KS" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "KY" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "LA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "ME" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "MD" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "MA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "MI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "MN" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "MS" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "MO" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "MT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "NE" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "NV" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "NH" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "NJ" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "NM" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "NY" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "NC" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "ND" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "OH" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "OK" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "OR" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "PA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "RI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "SC" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "SD" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "TN" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "TX" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "UT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "VT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "VA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "WA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "WV" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "WI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    Postal == "WY" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/", str_to_lower(State), ".csv"),
    TRUE ~ "https://the-art-of-data-journalism.github.io/tutorial-data/colleges/nebraska.csv"  # Default case if none match
  ))

colleges <- read_csv(stateName$dataurl)

collegerows <- nrow(colleges)
```
## The Goal

In this lesson, you'll learn how to create dumbbell charts, a powerful tool for visualizing the difference between two related values. By the end of this tutorial, you'll understand when to use dumbbell charts, how to prepare your data for this type of visualization, and how to create and customize dumbbell charts using ggplot2 and the `ggalt` package. You'll practice filtering data, reordering chart elements, and adding color to enhance the visual story. These skills will enable you to effectively communicate comparisons and gaps in your data, a crucial ability for data journalists looking to highlight meaningful differences.

## Why Visualize Data?

Coulter Jones has worked in tiny newspapers, trade publications and in public radio. He's also worked at two of the largest news organizations in the world -- Bloomberg News and the Wall Street Journal. In smaller places, it's much more common for someone with some data skills to take a graphic all the way to publication. But Bloomberg and the WSJ have graphics staffs larger than most newspapers these days. There's zero chance of Jones turning some data he has into a graphic that faces the public.

And yet, he makes graphics all the time.

"I would say almost every time I do an analysis, right?" he said. 

As a data analyst, one of the most important things you can do -- and one of the easiest once you get some practice -- is to convert your data into a picture. You learned this from Tukey. You learned this from Playfair. You learned this from Tufte. Data journalists do this all the time, and it's one of the best reasons to learn R. 

"One of the things that I did not appreciate at first, but then quickly learned to love about R is, oh, this is a visual platform too," Jones said. "So I can very easily just do a histogram. I can do a two factor line or point, sort of like show me age and sex or something like that.

"If you just chart it, you can sometimes just see the story immediately."

## The Basics

Second to my love of waffle charts because of their name, and I'm always hungry, dumbbell charts are an excellently named way of **showing the difference between two things on a number line** -- a start and a finish, for instance. Or the difference between two related things. When the gap between numbers is the news, dumbbell charts are what you want.

If you Google dumbbell charts, you'll find they come batteries included in a library called `ggalt`. The problem? `ggalt` has been abandoned, and is fading away slowly. It won't install easily now, which is a bad sign for the future. But the truth is, dumbbell charts are easy to make -- it's just a line and two dots. That's three geoms, and that's not hard. 

Let's give it a whirl.

```{r load-tidyverse, exercise=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

For this, let's use a list of a colleges from the Department of Education's College Scorecard. The dataset is massive -- there are nearly 6,500 colleges and universities in the dataset, which in the grand scheme of data isn't that many rows. But there's more than 3,300 *columns* of data. In other words, there's that many things about each college being tracked.

Let's focus on a couple of the most important things -- do people graduate? And what does it cost to go there?

We'll load the data for your state. I've cut down the number of columns to just what we need, and limited the the colleges to those granting associate and bachelor's degrees. Two other limitations: If a college didn't report data, they were dropped. And the number of colleges for a state is capped at 20 (the 20 largest by undergraduate enrollment). Dumbbell charts can only hold so many. 

```{r colleges-read, exercise=TRUE, exercise.reveal_solution = FALSE, exercise.setup = "load-data"}
colleges <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/colleges/____.csv")
```

```{r colleges-read-check}
grade_this({
  if (identical(nrow(.result), collegerows)) {
    pass("Great work! You imported colleges in your state.")
  }
  fail()
})
```

Let's look at what we've got here:

```{r head-data, exercise=TRUE, exercise.setup = "load-data"}
head(colleges)
```
```{r head-data-solution}
head(colleges)
```
```{r head-data-check}
grade_this_code()
```

For this example, let's look at the difference between a school's in-state tuition vs. out-of-state tuition. Most public colleges charge more for people who don't live there. The obvious reason is giving people who pay taxes in the state a break, and charging more for people who don't pay taxes there. Private schools don't have to do that, but some do. 

### Exercise 1: The first part of the dumbbell

Now, the way that the dumbbell works is pretty simple when viewed through what we've done before. There's just some tweaks. 

The first thing we need is a `geom_segment` -- a line segment. `geom_segment` takes four parts: where the y axis starts and ends and where the x axis starts and ends. 

First: We start with the y axis. The reason is we want our dumbbells going left and right so the label is going to be on the y axis. That's the column with the college name in it. Good news: where this starts and ends -- `y` and `yend` -- is the same in a dumbbell chart. We want our segment to stay on the line given to a particular college. 

Second: Our x is actually two things as well: `x` and `xend`. What you put in there will decide where on the line the segment starts and ends. What goes in the `x` and `xend` are the two numbers where the gap between them is meaningful. So for us, it's our two tuition columns, one for in-state and the second for out of state. 

```{r dumbbell2, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() + 
  geom_segment(
    data=colleges, 
    aes(
      y=____, yend = ____,
      x=____, xend=____
      )
  )
```
```{r dumbbell2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_segment(
    data=colleges, 
    aes(
      y=INSTNM, yend=INSTNM, 
      x=TUITIONFEE_IN, xend=TUITIONFEE_OUT
      )
  )
```
```{r dumbbell2-check}
grade_this_code()
```

Well, that's a chart alright. But we need dots -- the bells on our dumbbell -- and then we need to show which dot is the in-state tuition and what is the out of state. How? With colors.

### Exercise 2: Bells and colors

To make this work, we'll need to do three things: first, we're going to add two `geom_point`s. Those will be the bells in the the dumbbell. Our `geom_point`s will get two things: a color *inside* the aes and a size *outside* of it. Because the color goes *inside* the aes, we have to play with it a little. If we put a number column in there, the legend will give us a gradient instead of just a color. But if we put text in there, it will give us one color and a legend. The size is there to make the bells a little bigger than standard dot size. What goes in y stays the same as the `geom_segment`. What goes in `x` is what is in `x` and `xend` in your `geom_segment`. 

```{r dumbbell3, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() + 
  geom_segment(
    data=colleges, 
    aes(
      y=INSTNM, yend=INSTNM, 
      x=TUITIONFEE_IN, xend=TUITIONFEE_OUT
      )
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=____,
      x=____,
      color="In-state"
    ),
    size=2
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=____,
      x=____,
      color="Out-of-state"
    ),
    size=2
  )
```
```{r dumbbell3-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_segment(
    data=colleges, 
    aes(
      y=INSTNM, yend=INSTNM, 
      x=TUITIONFEE_IN, xend=TUITIONFEE_OUT
      )
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=INSTNM,
      x=TUITIONFEE_IN,
      color="In-state"
    ),
    size=2
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=INSTNM,
      x=TUITIONFEE_OUT,
      color="Out-of-state"
    ),
    size=2
  )
```
```{r dumbbell3-check}
grade_this_code()
```

And now we have a chart that is trying to tell a story. We know, logically, that the dot on the left is good, because it means cheaper tuition. A long distance between the two dots? That shows a gap between what in-state students pay and what out-of-state students pay. In some cases, that's small. In some it's *huge*. But what about the colleges that have just one color of dots? The issue there is that they don't have in-state and out-of-state tuition. They just have ... tuition. The dot is there -- it's just getting covered by the other dot.

### Exercise 3: Arrange helps tell the story

But what if we sort it by out-of-state tuition, so we see them in order of cost. 

```{r dumbbell4, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() + 
  geom_segment(
    data=colleges, 
    aes(
      y=reorder(____, ____), yend=INSTNM, 
      x=TUITIONFEE_IN, xend=TUITIONFEE_OUT
      )
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=reorder(____, ____),
      x=TUITIONFEE_IN,
      color="In-state"
    ),
    size=2
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=reorder(____, ____),
      x=TUITIONFEE_OUT,
      color="Out-of-state"
    ),
    size=2
  )
```
```{r dumbbell4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_segment(
    data=colleges, 
    aes(
      y=reorder(INSTNM, TUITIONFEE_OUT), yend=INSTNM, 
      x=TUITIONFEE_IN, xend=TUITIONFEE_OUT
      )
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=reorder(INSTNM, TUITIONFEE_OUT),
      x=TUITIONFEE_IN,
      color="In-state"
    ),
    size=2
  ) + 
  geom_point(
    data=colleges,
    aes(
      y=reorder(INSTNM, TUITIONFEE_OUT),
      x=TUITIONFEE_OUT,
      color="Out-of-state"
    ),
    size=2
  )
```
```{r dumbbell4-check}
grade_this_code()
```

```{r storysetup, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("Now we can start asking questions. What story is this telling? What is the most expensive place to go to college in {stateName$State}, at least among this list? What colleges have the widest gaps between in-state and out-of-state tuition?")
```

## The Recap

Throughout this lesson, you've learned how to create and customize dumbbell charts to visualize differences between two related values. You've practiced creating basic dumbbell charts, and enhancing them with color and size adjustments. You've also seen how reordering the chart elements can help tell a clearer story, as demonstrated with the comparison of in-state and out-of-state tuition costs across colleges. Remember, dumbbell charts are particularly effective when the gap between two values is the key story you want to tell. Consider how this visualization technique can help you highlight important comparisons in your datasets.

## Terms to Know

- Dumbbell chart: A type of chart that displays the difference between two related values for multiple categories, resembling a dumbbell shape.
- `ggalt`: An extension package for ggplot2 that includes additional geoms, including `geom_dumbbell()` for creating dumbbell charts.
- `geom_dumbbell()`: The specific function in ggalt used to create dumbbell charts in ggplot2.
- `x` and `xend`: In dumbbell charts, these aesthetics define the start and end points of each "dumbbell."
- `colour`, `colour_x`, `colour_xend`: Aesthetics in `geom_dumbbell()` used to set the colors of the connecting line and the start and end points of each dumbbell.
- `size`: An aesthetic in `geom_dumbbell()` used to adjust the thickness of the points and connecting line.
