---
title: "Data Journalism Lesson 23: Tables"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: journal
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to make a table with visual interestingness.
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(gt)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```
```{r load-data, message=FALSE, warning=FALSE}
state <- Sys.getenv("tutorial.state")
if(state == "") state <- "NE"

stateName <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/states.csv") |> filter(Postal == state) 

stateName <- stateName |> 
  mutate(dataurl = case_when(
    Postal == "AL" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "AK" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "AZ" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "AR" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "CA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "CO" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "CT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "DE" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "FL" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "GA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "HI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "ID" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "IL" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "IN" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "IA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "KS" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "KY" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "LA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "ME" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "MD" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "MA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "MI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "MN" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "MS" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "MO" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "MT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "NE" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "NV" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "NH" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "NJ" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "NM" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "NY" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "NC" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "ND" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "OH" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "OK" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "OR" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "PA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "RI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "SC" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "SD" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "TN" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "TX" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "UT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "VT" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "VA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "WA" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "WV" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "WI" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    Postal == "WY" ~ paste0("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/", str_to_lower(State), ".csv"),
    TRUE ~ "https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/nebraska.csv"  # Default case if none match
  ))

uninsured <- read_csv(stateName$dataurl)

uninsuredrows <- nrow(uninsured)

top10 <- uninsured |> 
  mutate(percent_uninsured = nui_pt/(nui_pt + nic_pt)) |> 
  top_n(10, wt=percent_uninsured)
```
## The Goal

In this lesson, you'll learn how to create informative and visually appealing tables using the gt package in R. By the end of this tutorial, you'll understand how to structure data for table creation, customize table elements such as headers and labels, and apply styling to enhance readability. You'll practice these skills using real-world data on estimates of people who have health insurance, gaining practical experience in presenting complex information in a clear, tabular format. This ability to create professional-looking tables will be invaluable for effectively communicating data findings in your journalism projects.

## Why Visualize Data?

There are few graphical forms of data more criticized than the data table. For a long time, they were all there was. And they were used heavily -- and not particularly creatively. And people knew it wasn't good.

"Information, that is imperfectly acquired, is generally as imperfectly retained; and a man who has carefully investigated a printed table, finds, when done, that he has only a very faint and partial idea of what he has read; and that like a figure imprinted on sand, is soon totally erased and defaced," Playfair wrote in 1786. 

Fast forward nearly 200 years, Tufte gives a little more room and grace to the table, but only under limited circumstances. And he has no love for pie charts.

"Tables are clearly the best way to show exact numerical values, although the entries can also be arranged in semi-graphical form," Tufte wrote. "Tables are preferable to graphics for many small data sets. A table is nearly always better than a dumb pie chart; the only worse design than a pie chart is several of them, for then the viewer is asked to compare quantities located in spatial disarray both within and between pies, as in this heavily encoded example from an atlas. Given their low data-density and failure to order numbers along a visual dimension, pie charts should never be used."

So a table is better than a pie chart, which should never be used. Got it. 

But now more than 40 years after Tufte first published The Visual Display of Quantitative Information, tables have evolved a bit further. Tables are still excellent at showing exact numerical values, but now we can incorporate more. Color, gradients, even graphics in each row. 

Let's explore.

## The Basics

A table is the most simple form of data presentation. Tables have been around for millenia. We've improved in them since, so we're not making a table. We're making a table with features. 

What features you ask? Because the library we're using today is different from ggplot, we're going to jump ahead a little bit and add required parts of every chart. There are some exceptions to this rule, but not many, and used only by people who really, really know what they're doing. The rest of us should live by every chart having these parts:

1. A headline.
2. Explanatory text under the headline, called chatter in the business.
3. A source line, where you say where you got the data from.
4. A credit line, where you put your name on the chart. All the glory and all the blame belong to you.

There are more parts that your chart might need -- annotations, arrows, shapes to indicate regions -- and we've talked about using color to draw a reader's eye, but every chart needs those four. So we're going to build a table with those four things ... and some color to draw the eye. 

Why a table? Sometimes, the best way to show your data is with a table -- simple rows and columns. It allows a reader to compare whatever they want to compare a little easier than a graph where you've chosen what to highlight. The folks that made R Studio and the Tidyverse have a neat package called `gt`.

For this assignment, we'll need `gt`.

What does this library do? We need the Tidyverse and gt to find out.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(gt)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(gt)
```
```{r load-tidyverse-check}
grade_this_code()
```

This is being written in the middle of a presidential election that seems like it never ended from the last one, and I have a terrible feeling that the next one will start as soon as this one is over. A permanent campaign issue seems to be health care -- who should get it, how it should be paid for, etc. It's such an issue that the Census Bureau has a program called the Small Area Health Insurance Estimates program, which ... you guessed it ... estimates the number of people who have health insurance in an area. It's an estimate, based on a model, but it's the best we've got if you've got a question about differences in your state when it comes to health insurance coverage. 

Let's take a look. Plug your state name in and load up the data:

```{r sahie-read, exercise=TRUE, exercise.reveal_solution = FALSE, exercise.setup = "load-data"}
uninsured <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/uninsured/____.csv")
```

```{r sahie-read-check}
grade_this({
  if (identical(nrow(.result), uninsuredrows)) {
    pass("Great work! You imported heath insurance estimates in your state.")
  }
  fail()
})
```

Tables, like bar charts, beeswarms, waffles and a few others, can only fit so many rows of data. Generally speaking, tables are great at top 10 lists. Even top 20 can work if you've got the space and your readers have the time. So let's first take a quick peek at our data and then we're going to make a list of the 10 counties with the largest percentage of uninsured people.

```{r head-data, exercise=TRUE, exercise.setup = "load-data"}
head(uninsured)
```
```{r head-data-solution}
head(uninsured)
```
```{r head-data-check}
grade_this_code()
```

The column `nic_pt` are the number of insured people and the column `nui_pt` are the number of uninsured people. One thing to realize about this data -- if you add up the insured people and the uninsured people, you will **not** get the number of people in the county. Why? The answer is pretty simple, once you think about it -- in the United States, Medicare kicks in for people 65 and over. If you dig into the SAHIE docs, they cut their data off at 64. What this all means is that the numbers we're looking at are the estimates of non-Medicare aged people. So something to keep in mind when wording certain things later. 

### Exercise 1: Making a top 10 list

To make a top 10 list, we have to convert our numbers into percentages first, then use top_n to get that list. In our mutate, we're going to calculate the percentage of *uninsured* people. 

<div class="alert alert-info">
<h4 class="alert-heading">Key Concept</h4> 
<p class="mb-0">Calculating a percentage is taking the small thing -- the component -- and dividing it by the total things. So if five students in a class of 20 are left-handed, the percentage is 5/20.</p>
</div>

We don't have the total number of insurable people, so we'll have to add our two insurance estimate numbers together to get it. We can use that new column in the `wt` of our `top_n` function to get the top 10. 

Then we'll name our new dataframe `top10`.

```{r table1, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
____ <- uninsured |> 
  mutate(percent_uninsured = ____/(nui_pt + nic_pt)) |> 
  top_n(10, wt=____)
```
```{r table1-solution, exercise.reveal_solution = FALSE}
top10 <- uninsured |> 
  mutate(percent_uninsured = nui_pt/(nui_pt + nic_pt)) |> 
  top_n(10, wt=percent_uninsured)
```
```{r table1-check}
grade_this_code()
```

Now, you can make a table with `gt` as simply as adding `gt()` after your dataframe. 

```{r table2, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> gt()
```
```{r table2-solution, exercise.reveal_solution = FALSE}
top10 |> gt()
```
```{r table2-check}
grade_this_code()
```

So there you have it. You made a table. You're done, right? Not hardly.

`gt` has a mountain of customization options. The good news is that it works in a very familiar pattern -- we start with data and we use |> to pipe to the next thing.

### Exercise 2: Limiting column names, fixing column names

We'll start with limiting the number of columns. You already know how to do this -- it's `select`. For this table, let's select the county name, the insured people, the uninsured people, and the percent uninsured in that order. While we're at it, let's add an arrange so we get them in order by percent uninsured.

Then, the header names for those columns are terrible. No one can read them beyond some very nerdy people (like us). We need to fix them. The way this works is we need to tell `cols_label` what columns we want to change and what we want to change them into. They go in that order -- column name from the data = "New Name I Want To Show People". 

```{r table3, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(____)) |> 
  gt() |> 
  cols_label(
    ____ = "County",
    ____ = "Insured people",
    ____ = "Uninsured people", 
    ____ = "Percent uninsured"
  )
```
```{r table3-solution, exercise.reveal_solution = FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  )
```
```{r table3-check}
grade_this_code()
```

Better. Note the pattern: Actual header name = "What we want to see". 

Now we can start working with styling. 

### Exercise 3: Styling, part 1

The truth is most of your code in tables is going to be dedicated to styling specific things. First we need: A headline and some chatter. They're required parts of a graphic so they're a good place to start. We do that with `tab_header` and in there we need a title and subtitle.

```{r table4, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(____)) |> 
  gt() |> 
  cols_label(
    ____ = "County",
    ____ = "Insured people",
    ____ = "Uninsured people", 
    ____ = "Percent uninsured"
  ) |>
  tab_header(
    ____ = "Where are the most uninsured people?",
    ____ = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  )
```
```{r table4-solution, exercise.reveal_solution = FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  )
```
```{r table4-check}
grade_this_code()
```

We have a headline and some chatter, but ... gross. Centered? The extra lines? No real difference in font weight? We can do better. 

### Exercise 4: Changing typography

We can style individual elements using `tab_style`. First, let's make the main headline -- the `title` -- bold and left aligned. Then, we can left align the `subtitle`, but leave it normal weight (i.e. not bold).

```{r table5, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(____)) |> 
  gt() |> 
  cols_label(
    ____ = "County",
    ____ = "Insured people",
    ____ = "Uninsured people", 
    ____ = "Percent uninsured"
  ) |>
  tab_header(
    ____ = "Where are the most uninsured people?",
    ____ = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("____")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("____")
  )
```
```{r table5-solution, exercise.reveal_solution = FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  )
```
```{r table5-check}
grade_this_code()
```

### Exercise 5: Adding source and credit lines

The next item on the required elements list: Source and credit lines. In `gt`, those are called a `tab_source_note`:

```{r table6, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(____)) |> 
  gt() |> 
  cols_label(
    ____ = "County",
    ____ = "Insured people",
    ____ = "Uninsured people", 
    ____ = "Percent uninsured"
  ) |>
  tab_header(
    ____ = "Where are the most uninsured people?",
    ____ = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("____")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("____")
  ) |>
  ____(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  )
```
```{r table6-solution, exercise.reveal_solution = FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  )
```
```{r table6-check}
grade_this_code()
```

## More styling

We can do a lot with `tab_style`. For instance, we can make the headers bold and reduce the size a bit to reduce font congestion in the area.

```{r table7, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) 
```
```{r table7-solution, exercise.reveal_solution = FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) 
```
```{r table7-check}
grade_this_code()
```

Next up: There's a lot of lines in this that don't need to be there. `gt` has some tools to get rid of them easily and add in some other readability improvements: `opt_row_striping` and `opt_table_lines`.

More magic that gt has? You can automatically format percentages without having to go back to your original data. We'll add that below the opts.

```{r table8, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   )  |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
  fmt_percent(
    columns = c(percent_uninsured),
    decimals = 1
  )
```
```{r table8-solution, exercise.reveal_solution = FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
  fmt_percent(
    columns = c(percent_uninsured),
    decimals = 1
  )
```
```{r table8-check}
grade_this_code()
```

Throughout this series, we've been using color and other signals to highlight things. With a little `tab_style` magic, we can change individual rows and add color. The last `tab_style` block here will first pass off the styles we want to use -- we're going to make the rows red and the text white -- and then for locations we specify where with a simple filter. What that means is that any rows we can address with logic -- all rows with a value greater than X, for example. Or the largest number in our data. Let's use that. We'll use the `max()` function and highlight the county with the largest single number of uninsured people in the top 10.  

```{r table9, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
  fmt_percent(
    columns = c(percent_uninsured),
    decimals = 1
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "red"),
      cell_text(color = "white")
      ),
    locations = cells_body(
      rows = nui_pt == max(nui_pt))
  )
```
```{r table9-solution, exercise.reveal_solution = FALSE}
top10 |> 
  select(ctyname, nic_pt, nui_pt, percent_uninsured) |> 
  arrange(desc(percent_uninsured)) |>   
  gt() |> 
  cols_label(
    ctyname = "County",
    nic_pt = "Insured people",
    nui_pt = "Uninsured people", 
    percent_uninsured = "Percent uninsured"
  ) |>
  tab_header(
    title = "Where are the most uninsured people?",
    subtitle = "Here are the top counties in the state by uninsured percentage, according to the Census Bureau."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Your Name  |  **Source:** US Census Bureau")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
  fmt_percent(
    columns = c(percent_uninsured),
    decimals = 1
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "red"),
      cell_text(color = "white")
      ),
    locations = cells_body(
      rows = nui_pt == max(nui_pt))
  )
```
```{r table9-check}
grade_this_code()
```

We've arrived where we want to be: We've created a clear table that allows a reader to compare counties at will while also using color to draw attention to the thing we want to draw attention to. **We've kept it simple so the color has impact**.

Note for tables: Copy and paste is your friend. This *looks* like a lot of code, but 90 percent of it is copy, paste, edit slightly and move on. Once you develop a style on tables, you'll copy and paste that over and over again and adjust for your new column names. Getting your first one done is a lot of work. Getting your second one done takes minutes. 

## The Recap

Throughout this lesson, you've mastered the art of creating sophisticated tables using gt. You've learned to transform raw data into structured, easy-to-read tables by customizing headers, applying styles, and highlighting key information. Remember, effective tables strike a balance between informativeness and visual appeal. You'll find that well-designed tables can be powerful tools for presenting complex data in a format that's accessible to your audience. Keep experimenting with different styling options to find the perfect balance for your specific data and story needs.

## Terms to Know

- `gt`: An R package designed for creating publication-quality tables.
- `tab_header()`: A `gt` function used to add a title and subtitle to your table.
- `cols_label()`: A `gt` function used to rename column headers in your table.
- `tab_style()`: A `gt` function that allows you to apply custom styles to specific parts of your table.
- `tab_source_note()`: A `gt` function used to add source information or notes at the bottom of your table.
