---
title: "Data Journalism Lesson 18: Slope charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: journal
    ace_theme: github
runtime: shiny_prerendered
description: >
  Comparing one time to another one
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(glue)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```
```{r load-data, message=FALSE, warning=FALSE}
state <- Sys.getenv("tutorial.state")

stateName <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/states.csv") |> filter(Postal == state) 

peak <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/peak-unemployment/peak.csv")

state_peak <- peak |> filter(Location == stateName$State)

staterows <- nrow(state_peak)
```

## The Goal

In this lesson, you'll learn how to create slope charts, a powerful tool for visualizing changes between two time periods. By the end of this tutorial, you'll understand when to use slope charts, how to prepare your data, and how to construct these charts using ggplot. You'll practice filtering data, manipulating chart elements, and adding informative labels to create clear and impactful visualizations. These skills will enable you to effectively communicate trends and comparisons in your data journalism projects, especially when dealing with before-and-after scenarios or year-over-year changes.

## Why Visualize Data?

More to come.

## The Basics

A line chart, as we learned, shows change over time, with the date on the x-axis and each line representing a state or county or some other entity. But sometimes, you have just two time periods -- a line chart of two years is ... not a line chart. 

But with some fiddling with lines and points, you can create a new chart type that does show change over two time periods quite well: A slope chart. 

Think of a slope chart like the meme you see online: How is started vs how it's going. 

For purposes of this exercise, we're going to look at unemployment again, but this time, we're just looking at April 2020 and April 2024. April 2020 is the month when unemployment was higest for almost every state. How does that compare four years later? 

Think of it like playing connect the dots from elementary school.

The difference between slope charts and most other chart types in ggplot is that there is no geom for a slope chart. We're going to make one using pieces and parts of other geoms. Combine the parts together and you get a chart. This is another example of where layering is going to come in very handy.

We'll need the tidyverse.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And we'll grab the data, which is just one file for the nation. 


```{r load-data-exercise, exercise = TRUE}
peak <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/peak-unemployment/peak.csv")
```
```{r load-data-exercise-solution}
peak <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/peak-unemployment/peak.csv")
```
```{r load-data-exercise-check}
grade_this_code()
```

We should take a quick look at our data so we know what we're working with. We'll use `head` for that. 

```{r head-data, exercise=TRUE, exercise.setup = "load-data"}
head(peak)
```
```{r head-data-solution}
head(peak)
```
```{r head-data-check}
grade_this_code()
```

### Exercise 1: Our first slope chart

Our data is in pretty good shape already, so we can go right to making a chart. 

A slope chart is really made up of two separate geoms that combine to make a form -- a geom_point and a geom_line. The parts inside the two geoms take is identical. In this case, we're going to have the x be the `year` because like a line chart, our x has to be a date or date-like thing. The y is our unemployment rate column and we're going to add a group value, which is the `Location`. I'm also going to go ahead and set the color for these to grey, which you'll see why in a moment. 

```{r slope2, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() + 
  geom_line(data=____, aes(x=____, y=____, group=____), color="grey") +
  geom_point(data=____, aes(x=____, y=____, group=____), color="grey")
```
```{r slope2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_point(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey")
```
```{r slope2-check}
grade_this_code()
```

Well, we got something. But who changed the most? What should we focus on? We don't have anything to guide us here.  

```{r storysetup, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("Might as well be selfish and focus on {stateName$State}. Unless you work for a major national news organization, you're going to be focusing on where you live -- on the audience you serve. People care about where they are, so give them what they want.")
```

### Exercise 2: Filtering

To let us focus, let's build a dataframe called `state_peak` and filter the `Location` by your state name.

```{r peak-read, exercise=TRUE, exercise.reveal_solution = FALSE, exercise.setup = "load-data"}
state_peak <- peak |> 
  filter(Location == "____")
```

```{r peak-read-check}
grade_this({
  if (identical(nrow(.result), staterows)) {
    pass("Great work! You've got it.")
  }
  fail()
})
```

Now, with a state to focus on, we can set about making two new geoms -- identical to our previous ones, just with a different dataframe and a different color.

### Exercise 3: Give readers a focus

```{r slope3, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() + 
  geom_line(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_point(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_line(data=____, aes(x=____, y=____, group=____), color="red") +
  geom_point(data=____, aes(x=____, y=____, group=____), color="red")
```
```{r slope3-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_point(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_line(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") +
  geom_point(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red")
```
```{r slope3-check}
grade_this_code()
```

Now we can start with our questions: What story does this tell? Is it very clear? 

### Exercise 4: Getting creative with spacing and breaks

Now that we have something to work with, it's time to start improving it. It would be good to label the state we're highlighting. Maybe we label them on each end of the graph to help people follow the line. Given how much space the default ggplot behavior has left us to do that, we're going to need to make some space.

We can do that with some clever manipulations of the x axis. In ggplot, there are all kinds of ways to work with the x and y axis scales. In this case, we have a year on the x axis, which to R is just a number, so we'll use `scale_x_continuous` to work with that kind of scale. We're going to first set the line breaks manually so there's just 2020 and 2024. Then, we're going to set the limits of our x axis so that it doesn't just default to 2020-2024, but is a little wider so we have extra space. We're going to add one year on each side to pad it out. But think about it -- how do you *add* one year of space to the *left* of 2020? You subtract one year. How do you *add* a year of space to the *right* of 2024? You add one year. 

```{r slope4, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() + 
  geom_line(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_point(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_line(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") +
  geom_point(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") + 
  scale_x_continuous(breaks = c(____, ____), limits = c(____, ____))  
```
```{r slope4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_point(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_line(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") +
  geom_point(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") + 
  scale_x_continuous(breaks = c(2020, 2024), limits = c(2019, 2025)) 
```
```{r slope4-check}
grade_this_code()
```

A ha! Better. We now just have the two years and we have some space to put our state name in there to help people.

### Exercise 6: adding labels

To add the labels, we again need to be a bit clever. We want to label each dot *twice* -- once for 2020 and once for 2024. To do this *looks* like a lot, but it really isn't. It's two rows for each state, but most of it is copy and paste work. Once you get one state working, you can add more with copy, paste, change the dataframe and done. 

The first problem we have is our `state_peak` data has both years in it, so we're going to have to do something we don't do a lot of: put a `dplyr` filter into our data= bits. 

We're going to use a `geom_text`, which will just put text anywhere on your graph. The aes will be the same as the geom_point, except it will have a label in it that we'll set to the `Location`. Then, for data, we're going to filter our dataframes for the `min(year)` and the `max(year)`, which will give us 2020 and 2024 respectively. Then, one last little trick -- we're going to nudge our labels over a little. For our min, we're going to subtract .5 years off our `year`, which will have the effect of moving the text to the left just a bit. Then, on the max side, we'll nudge it .5 years to the right by adding it to the `year`. How did I come to .5? The truth is I just found it through trial and error. Tried something, adjusted from there. Rinse and repeat until I got the number I liked.

```{r slope5, exercise=TRUE, exercise.setup = "load-data", message=FALSE}
ggplot() + 
  geom_line(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_point(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_line(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") +
  geom_point(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") + 
  geom_text(data=____ |> filter(year == max(____)), aes(x=year + .5, y=Unemployment_Rate, group=Location, label=____)) + 
  geom_text(data=____ |> filter(year == min(____)), aes(x=year - .5, y=Unemployment_Rate, group=Location, label=____)) + 
  scale_x_continuous(breaks = c(2020, 2024), limits = c(2019, 2025)) 
```
```{r slope5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_point(data=peak, aes(x=year, y=Unemployment_Rate, group=Location), color="grey") +
  geom_line(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") +
  geom_point(data=state_peak, aes(x=year, y=Unemployment_Rate, group=Location), color="red") + 
  geom_text(data=state_peak |> filter(year == max(year)), aes(x=year + .5, y=Unemployment_Rate, group=Location, label=Location)) + 
  geom_text(data=state_peak |> filter(year == min(year)), aes(x=year - .5, y=Unemployment_Rate, group=Location, label=Location)) + 
  scale_x_continuous(breaks = c(2020, 2024), limits = c(2019, 2025)) 
```
```{r slope5-check}
grade_this_code()
```

And there you have it. A clean and clear slope chart showing how the peak of the pandemic compares to four years later. 

## The Recap

Throughout this lesson, you've mastered the art of creating slope charts to visualize changes between two time periods. You've learned how to prepare your data, construct the basic chart structure, and enhance it with colors, labels, and spacing. Remember, slope charts are particularly effective for showing how different entities (like states in our example) change relative to each other over time. As you go forward, you'll find slope charts to be a valuable addition to your visualization toolkit, especially when you need to clearly communicate before-and-after comparisons or year-over-year changes in a compelling and easy-to-understand format.

## Terms to Know

- Slope chart: A type of chart that shows changes between two time periods by connecting data points with lines.
- `geom_line`: A ggplot2 function used to create lines in a chart, essential for connecting points in a slope chart.
- `geom_point`: A ggplot2 function used to create points in a chart, used to mark the start and end points in a slope chart.
- `geom_text`: A ggplot2 function used to add text labels to a chart, often used to label data points in slope charts.
- `scale_x_continuous`: A ggplot2 function used to customize the x-axis when working with numerical (i.e. continuous) data.
