---
title: "Data Journalism Lesson 20: Scatterplots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: journal
    ace_theme: github
runtime: shiny_prerendered
description: >
  Comparing two numbers to uncover a trend.
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(glue)
library(cli)
library(ggrepel)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```
```{r load-data, message=FALSE, warning=FALSE}
state <- Sys.getenv("tutorial.state")
if(state == "") state <- "NE"

stateName <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/states.csv") |> filter(Postal == state) 

colleges <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/college-cost/national.csv")

state_colleges <- colleges |> filter(STABBR == state)

state_colleges_count <- nrow(state_colleges)

collegeplural <- pluralize("There {?is/are} {state_colleges_count} {?college/colleges} in the state.")

collegeplural2 <- pluralize("{state_colleges_count} institution{?s} that grant bachelor's degrees.")
```
## The Goal

In this lesson, you'll learn how to create and interpret scatterplots, a powerful tool for visualizing relationships between two numerical variables. By the end of this tutorial, you'll understand how to prepare data for scatterplots, construct basic plots using ggplot2, highlight specific data points, and add labels using ggrepel. You'll practice these skills using real-world college completion and cost data, gaining insight into how to uncover and communicate trends and outliers in your datasets. These abilities will be crucial for identifying patterns and relationships in your data journalism projects.

## Why Visualize Data?

One of the more interesting thinkers about data visualization on planet earth today is Alberto Cairo. He is the Knight chair in visual journalism at the School of Communication at the University of Miami. He's written four books in English about data visualization -- The Functional Art (2013), the Truthful Art (2016), How Charts Lie (2019) and The Art of Insight (2023). If you're serious about data visualization, you own these books and if he is speaking somewhere near you, you go and you listen. 

I was hooked early by Cairo's vision for data visualization -- and the kind of person who does the work -- from the first chapter of The Functional Art. Specifically, he wrote that "the life of a visual communicator should be one of systematic and exciting intellectual chaos." I felt like my own read-anything-that-catches-my-attention lifestyle was all okay now. Intellectual curiosity -- spread far and wide -- leads to inspiration, which leads to all kinds of interesting things. 

For me -- your humble author -- Tufte and Cairo are two voices in my head when I am trying to make something out of data. They're fighting for the soul of my idea. On the one side, there is comfort in rules that simplify and dictate forms and functions. On the other is Cairo, talking about art and creativity and how rules can squash something great into something mediocre. 

What do I mean by this? And what do I mean by Cairo being one of the most fascinating thinkers about data visualization?

In August 2023, Cairo made an appearance on the Real World Data Science podcast to talk about his not-yet-released book The Art of Insight. In it, he talked about how visualization is like language in that it has dialects. The dialects of data viz are the places that use them -- journalism, science, industry, art, and others. The rules for how we write in journalism, for example, are vastly different from how we write in the sciences. So why are we trying to create rigid rules for all data visualization? 

"There are really no universal rules in data visualization: they are parochial rules that are applicable to the different dialects," Cairo said. "But it is wrong --
it is a mistake -- to apply the standards of one of the dialects of data visualization to a completely different dialect of data visualization." 

And, not just that, the "rules" are kind of a mishmash of empirical research and opinions. 

"There are a few things that could be considered rules," he said. "For example we know that if you want to compare numbers a bar graph is usually superior to a pie chart, for example. We know that there's empirical evidence behind that so you can sort of like derive a principle out of that, right, but beyond those very basic things there are really not many rules."

What do we have? Guidelines. Inherited wisdom. Historical examples. And you should learn them, Cairo said, along with the perception and cognition research that can guide us. 

"But after you do that all that matters is the choices that you make with the knowledge that you have and with the guesses that you can make," he said.

Today we're working with what I believe is the most versatile and interpretive graph type around -- the scatterplot. What makes the scatterplot so great? You use it to compare two dimensions of data -- so your x and y are both numbers. That means the space and the arrangement of dots in it is meaningful. If they cluster together? That means something. If they arrange along a path going up and to the right? Meaningful. Down and right? Meaningful. You can even turn the space into quadrants -- above and below average on your two dimensions of data -- and where dots appear means something. And, unlike many other graphical types, scatterplots are better, in my opinion, when there is *more* data. A scatterplot with just a few points isn't a very compelling scatterplot. A scatterplot with *thousands* can be.  

## The Basics

One of the most powerful things that we can do with visualizing data is establish relationships and find when things don't fit the relationship. 

In the last assignment, we looked at colleges, at the gap between in-state and out-of-state tuition. Next question: Is there a relationship between how much it costs to go to a college and how many people graduate? In other words -- if it's really expensive to go there, is that creating problems for people finishing their degrees? Or is it that the more expensive the school, the more likely those people graduate? Maybe they only let smart people in. Maybe they only let rich people in who don't worry about the cost. Maybe all of those things. We don't know. 

We'll start with libraries. We're adding `ggrepel` to this one because one common thing to do is label dots in the scatterplot. `ggrepel` does that. 

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(ggrepel)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(ggrepel)
```
```{r load-tidyverse-check}
grade_this_code()
```

We're going to work with a slightly different version of the college data. In this case, the differences are: 

* We're only looking at four-year colleges, ones that primarily grant bachelor's and graduate degrees, for the entire nation.
* The two fields we're going to look at are the average cost of attendance and the completion rate. The average cost of attendance -- `COSTT4_A` in the data -- is the total cost for a year at that school, including tuition, housing, books and fees. The completion rate -- `C150_4` in the data -- is the percent of first-time, first-semester freshmen who complete their degree in *6* years.
* I have filtered out schools who didn't report one or both of these bits of data. Example: There's 23 schools who reported zero students graduating in 6 years. That's ... odd.

Let's import the data now.

```{r load-data-exercise, exercise = TRUE}
colleges <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/college-cost/national.csv")
```
```{r load-data-exercise-solution}
colleges <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/college-cost/national.csv")
```
```{r load-data-exercise-check}
grade_this_code()
```

Scatterplots are, for my money, the most revealing and useful chart type out there that isn't a map. They reveal so much, and they allow you to make comparisons between two numbers. With a scatterplot, we put what we think predicts the thing on the x axis, and the thing being predicted on the y axis. `geom_point` works just like other geoms -- we need an x and a y and a data element. Both x and y are numbers. 

The other advantage of scatterplots? You can put a *huge* amount of data on them. There are more than 1,700 colleges that fit our criteria. Do we need to narrow it down for us to fit them all in the chart? Nope.

Let's take a look at our data so we know what we can work with going forward. 

```{r scatter2, exercise=TRUE, exercise.setup = "load-data"}
head(colleges)
```
```{r scatter2-solution, exercise.reveal_solution = FALSE}
head(colleges)
```
```{r scatter2-check}
grade_this_code()
```

### Excercise 1: Making a scatterplot

Remember -- we're looking at how much it costs to go to that school and comparing it to the completion rate. We talked about exactly which of the columns you have are those two things above.

```{r scatter3, exercise=TRUE, exercise.setup = "load-data"}
ggplot() + 
  geom_point(data=colleges, aes(x=____, y=____), color="grey")
```
```{r scatter3-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=colleges, aes(x=COSTT4_A, y=C150_4), color="grey")
```
```{r scatter3-check}
grade_this_code()
```

The way to read this chart -- when you're looking for relationships -- is do the dots sort of cluster in a pattern? In this case ... they do! At the lower costs, they all kind of bunch together, but the more expensive you get, the higher the completion percentage gets. See how they flow up and to the right? That's the relationship. The more expensive the school, the more likely a student is to complete it. A massive question that should be rattling around in your head right now: Why?

```{r storysetup, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("Where are {stateName$State} colleges on this chart? {collegeplural} You may have noticed we set the others to grey. Let's make these a color. But first, we have to make a dataframe of them.")
```

### Exercise 2: Colleges in your state 

The way to highlight a dot is to make a dataframe of one and add another geom_point. If you change the color of that one dot, it stands out. First, the dataframe. We'll call it `state_colleges` and the state abbreviation you want is your state's two-letter postal code. 

```{r college-read, exercise=TRUE, exercise.reveal_solution = FALSE, exercise.setup = "load-data"}
____ <- colleges |> filter(STABBR == "____")
```

```{r college-read-check}
grade_this({
  if (identical(nrow(.result), state_colleges_count)) {
    pass("Great work! You now have your state colleges.")
  }
  fail()
})
```

```{r storysetup2, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("Now we have a dataframe of {state_colleges_count}. Let's use it in a new geom.")
```

### Exercise 3: Red dots please

```{r scatter5, exercise=TRUE, exercise.setup = "load-data"}
ggplot() + 
  geom_point(data=colleges, aes(x=____, y=____), color="grey") + 
  geom_point(data=____, aes(x=____, y=____), color="red")
```
```{r scatter5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=colleges, aes(x=COSTT4_A, y=C150_4), color="grey") +
  geom_point(data=state_colleges, aes(x=COSTT4_A, y=C150_4), color="red")
```
```{r scatter5-check}
grade_this_code()
```

```{r storysetup3, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("And there's a chart of {stateName$State}'s {collegeplural2} But who is who?")
```

To get labels on there, we're going to add another geom. This one is called `geom_text_repel` and it's going to add text near each dot -- a `label` -- without them overlapping. The good news -- your `geom_text_repel` aes is exactly the same as your geom_point aes, just adding a `label`. And what column would make the most sense here to label the dot? 

### Exercise 4: Labeling

```{r scatter6, exercise=TRUE, exercise.setup = "load-data"}
ggplot() + 
  geom_point(data=colleges, aes(x=____, y=____), color="grey") + 
  geom_point(data=____, aes(x=____, y=____), color="red") + 
  geom_text_repel(data=____, aes(x=____, y=____, label=____))
```
```{r scatter6-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=colleges, aes(x=COSTT4_A, y=C150_4), color="grey") +
  geom_point(data=state_colleges, aes(x=COSTT4_A, y=C150_4), color="red") + 
  geom_text_repel(data=state_colleges, aes(x=COSTT4_A, y=C150_4, label=INSTNM))
```
```{r scatter6-check}
grade_this_code()
```

```{r storysetup4, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
if (state_colleges_count <= 5) {
  glue("And now you know which and where {stateName$State}'s colleges reside in this Cheap/Expensive vs. Graduates Nobody/Everybody comparison.")
} else {
  glue("And now you know which and where {stateName$State}'s colleges reside in this Cheap/Expensive vs. Graduates Nobody/Everybody comparison. But let's be honest: is it kind of a mess? Do you even see all of them? Do you need that many labels on there? What you would normally do at this point is pick and choose who you label. Is is just the largest? Is it a specific set of schools? You probably don't want more than about five to avoid cluttering up your scatterplot.")
}
```

## The Recap

Throughout this lesson, you've mastered the basics of creating informative scatterplots. You've learned how to plot multiple data points to reveal overall trends, highlight specific points of interest, and add labels to key data points. Remember, scatterplots are excellent for visualizing relationships between two variables and identifying outliers. As you gain more experience, you'll find scatterplots invaluable for uncovering stories within your data, such as the relationship between college costs and completion. Always consider how you can use color, labels, and additional data points to enhance the story your scatterplot tells.

## Terms to Know

- Scatterplot: A type of chart that shows the relationship between two numerical variables, with each data point represented as a dot on a two-dimensional plane.
- `ggrepel`: An R package that provides text and label geoms for ggplot2 that repel each other to avoid overlapping.
- `geom_point()`: A ggplot2 function used to create the points (dots) in a scatterplot.
- `geom_text_repel()`: A function from the ggrepel package used to add non-overlapping text labels to points in a plot.
- `aes()`: Short for "aesthetics," this ggplot2 function maps variables in your data to visual properties in the plot.
- `x-axis`: The horizontal axis in a scatterplot, typically representing the independent variable.
- `y-axis`: The vertical axis in a scatterplot, typically representing the dependent variable.
