---
title: "Data Journalism Lesson 24: The importance of text"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: journal
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to add important labels and annotations to your charts.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(glue)
library(tidyverse)
library(ggrepel)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

In this lesson, you'll learn about the critical role of text elements in creating effective data visualizations for journalism. By the end of this tutorial, you'll understand how to craft compelling headlines, write informative subtitles, add clear axis labels, and use strategic annotations to guide readers through your charts. You'll practice applying these text elements to a scatterplot, gaining hands-on experience in transforming raw data into a story-driven visualization. These skills will enable you to create more impactful and informative graphics that effectively communicate your data stories to readers.

## Why Visualize Data

In 2016, a group of Harvard and MIT researchers set out to answer a set of what seems like simple questions: What gets people's attention when looking at a chart? And what do they remember? 

The answer surprises people who have never made charts before. Because the answer is, at least at first, the headline on the chart.

That's right: Words. Not shapes or colors. Words. When you track people's eyes, that's where they go first, more often than not. 

"When participants were shown visualizations with titles during encoding, the titles were fixated 59% of the time," according to [Borkin et. al](http://olivalab.mit.edu/Papers/07192646.pdf). "Correspondingly, during the recall phase of the experiment, titles were the element most likely
to be described if present."

So not only is the title the thing people are looking at first, it's the title that acts as an anchor to memory. 

In short: the headline (or title depending on who you are asking) is incredibly important. In video storytelling, there's an old saying: If you have bad audio, you have bad video. A similar saying here could be this: if you have a bad title, you have a bad graphic.

In fact, [researchers at the University of Illinois in 2018](https://www.zcliu.org/vistitles/CHI18-VisTitles.pdf) found that if you give people a chart, and you give one group a title that frames it positively and another group that frames it negatively, their interpretation of the chart will be influenced by the headline. In other words, give people a graph showing the volume of water is 50 percent. One is given a title that says something to the effect of half full, the other gets half empty, and both will walk away with very different feelings about exactly the same data.

Designers have always known that text is important, they just never had the science to back them up. No less than William Playfair in 1801 wrote "that those who do not, at the first sight, understand the manner of inspecting the Charts, will read with attention the few lines of directions facing the first Chart, after which they will find all the difficulty entirely vanish, and as much information may be obtained in five minutes as would require whole days to imprint on the memory."

Just be careful what you're imprinting.

## The Basics

We started down this road a bit in the tables tutorial, but repeating some important stuff from that, text is a vital part of a good graphic. Some text is absolutely required. 

These are the pieces of a good graphic:

-   Headline
-   Chatter
-   The main body
-   Annotations
-   Labels
-   Source line
-   Credit line

The first on that list is the first for a reason. The headline is an incredibly important part of any graphic: it's often the first thing a reader will see. It's got to entice people in, tell them a little bit about what they're going to see, and help tell the story.

The second item is the chatter -- the text underneath that headline. It needs to work with the headline to further the story, drive people toward the point, maybe add some context.

The two bits of text are extremely important. Let's set up a chart and talk about how to do it wrong and how to do it better.

```{r load-tidyverse, exercise=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
```

```{r load-tidyverse-solution}
library(tidyverse)
library(ggrepel)
```

```{r load-tidyverse-check}
grade_this_code()
```

```{r text-load-data, message=FALSE, warning=FALSE}
national <- read_csv("https://the-art-of-data-journalism.github.io/tutorial-data/college-cost/national.csv")

exclusivencheap <- national |> 
  filter(ADM_RATE < .2) |> 
  filter(COSTT4_A < 50000)

natstats <- national |> 
  summarize(
    average_admit_rate = round(mean(ADM_RATE)*100, 0),
    average_cost = round(mean(COSTT4_A),0)
  )
```

The data and the chart code isn't important for you to follow along. The code is nothing special -- it's a riff off of work we've done before. Under the hood, I've loaded up all colleges and universities, and we're going to look at comparing how exclusive a school is vs how much it costs.

Let's bring up the scatterplot with an odd group of schools -- colleges that admit fewer than 20 percent of applicants, but cost less than $50,000 a year to attend -- as a starting place.

```{r bubble1, exercise=TRUE, exercise.setup = "text-load-data", message=FALSE}
ggplot() + 
  geom_point(
    data=national, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="grey"
    ) + 
  geom_point(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="red"
    ) + 
  geom_text_repel(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A, label=INSTNM),
    )
```
```{r bubble1-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(
    data=national, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="grey"
    ) + 
  geom_point(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="red"
    ) + 
  geom_text_repel(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A, label=INSTNM),
    )
```
```{r bubble1-check}
grade_this_code()
```

This chart is missing major parts -- required for any chart.

To fix this, we add labels -- `labs`. Each graphic -- regardless of type -- needs a title, subtitle, caption, x and y. Your title is your headline. The subtitle is called chatter -- a sentence under the headline that explains a little about the graphic. The caption is where you'll put the source of your data and the credit line -- your name. x and y should be in each `labs`, but sometimes what x and y are is obvious and you can leave them blank. For example: A bar chart with states on the y axis doesn't need to label them as States. It's obvious by the rest of the chart. In our case, nothing is obvious unless you've seen this data before. None of our readers have.

First, let's start with some headline basics:

-   Your headline should be about what the chart is about, **not what makes up the chart**. What story is the chart telling? What made it interesting to you? Don't tell me what the stats are, tell me what it says.
-   Your headline should be specific. **Generic headlines are boring and ignored**.
-   Your headline should, most often, have a verb. It's not a 100 percent requirement, but a headline without a verb means you're trying to be cute and ...
-   Your headline shouldn't be overly cute. Trying to get away with slang, a very Of The Moment cultural reference that will be forgotten soon, or some inside joke is asking for trouble.
-   Your headline should provoke a reaction.

### Exercise 1: Adding a headline, some chatter and some better labels

Let's add some text to our labs here. For your x and y labels, spell out and properly space and capitalize the initial characters of the column names. For a headline, use `Exclusive doesn't mean expensive` and for a subtitle, use this: `According to the federal Department of Education, there are seven schools who admit fewer than 20 percent of applicants but are among the most affordable.` On our x axis is the `Admit rate` and on the y axis is the `Average annual cost`.

```{r bubble2, exercise=TRUE, exercise.setup = "text-load-data", message=FALSE}
ggplot() + 
  geom_point(
    data=national, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="grey"
    ) + 
  geom_point(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="red"
    ) + 
  geom_text_repel(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A, label=INSTNM),
    ) + 
  labs(
    x="____", 
    y="____", 
    title="____", 
    subtitle="____", 
    caption="Source: US Department of Education | By Your Name"
    )
```

```{r bubble2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(
    data=national, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="grey"
    ) + 
  geom_point(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="red"
    ) + 
  geom_text_repel(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A, label=INSTNM),
    ) + 
  labs(
    x="Admit rate", 
    y="Average annual cost", 
    title="Exclusive doesn't mean expensive", 
    subtitle="According to the federal Department of Education, there are seven schools who admit fewer than 20 percent of applicants but are among the most affordable.", 
    caption="Source: US Department of Education | By Your Name"
    )
```
```{r bubble2-check}
grade_this_code()
```

The headline here works because it lays out a provocative premise (especially for parents). The chatter reveals what's in the chart without pointing at it -- show don't tell. Our x and y labels are clearer and let the reader know what's going on. 

## Annotations

Another important part of a chart -- but not every chart -- is annotation. Sometimes, you need to label something in a chart to help the reader out. It can be a point or series of points. Or it can be regions of a chart. Let's return to our bubble chart.

Annotations also help us draw attention to things, or help the reader understand what they're looking at.

We're going to add some lines to our chart that represent the average admit rate and average cost. To get that, we first have to calculate that. It's simple -- just a summarize and two means.

### Exercise 2: The means

If you don't remember what these columns are, look at your chart code above. They're the x and y in the bubble chart.

```{r bubble3, exercise=TRUE, exercise.setup = "text-load-data", message=FALSE}
national |> 
  summarize(
    average_admit_rate = mean(____),
    average_cost = mean(____)
  )
```

```{r bubble3-solution, exercise.reveal_solution = FALSE}
national |> 
  summarize(
    average_admit_rate = mean(ADM_RATE),
    average_cost = mean(COSTT4_A)
  )
```

```{r bubble3-check}
grade_this_code()
```

```{r storysetup, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "text-load-data", results='asis'}
glue("So what does that say? It says the average college admits about {natstats$average_admit_rate} percent of students and costs an average of ${natstats$average_cost} to attend for a year.")
```

To add lines for the average admit rate and average cost, we're going to use `geom_hline` and `geom_vline` which mean horizontal line and vertical line. The averages will create lines to divide the chart into four corners. On the upper right quadrant -- lots of kids getting admitted, super expensive. On the lower left quadrant, nobody getting in, but super cheap. The geoms just require an intercept -- `xintercept` for `geom_vline` and `yintercept` for the `geom_hline`. We get the values for those from the averages we just made.

To label them, I'm going to use geom_text and I'm just going to put numbers in the x and the y to move them where I need them. I can get those numbers by looking at the x and y axis of my chart and guessing. After that, it's increasing or decreasing the number, depending on which direction I want to move it around.

### Exercise 3: Adding lines and labels

Note here -- I'm only labeling two sections of the chart. There is such a thing as too much. We want to tell the story here with the most efficient use of elements -- text, color, shape.

```{r bubble4, exercise=TRUE, exercise.setup = "text-load-data", message=FALSE}
ggplot() + 
  geom_point(
    data=national, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="grey"
    ) + 
  geom_point(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="red"
    ) + 
  geom_text_repel(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A, label=INSTNM),
    size=3
    ) + 
  geom_vline(xintercept = ____) + 
  geom_hline(yintercept = ____) +
  geom_text(aes(x=.87, y=85000, label="High admits, high cost"), color="blue") +
  geom_text(aes(x=.3, y=15000, label="Fewer admits, below average cost"), color="blue") + 
  labs(
    x="Admit rate", 
    y="Average annual cost", 
    title="Exclusive doesn't mean expensive", 
    subtitle="According to the federal Department of Education, there are seven schools who admit fewer than 20 percent of applicants but are among the most affordable.", 
    caption="Source: US Department of Education | By Your Name"
    )
```

```{r bubble4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(
    data=national, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="grey"
    ) + 
  geom_point(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A),
    color="red"
    ) + 
  geom_text_repel(
    data=exclusivencheap, 
    aes(x=ADM_RATE, y=COSTT4_A, label=INSTNM),
    size=3
    ) + 
  geom_vline(xintercept = 0.7246474) + 
  geom_hline(yintercept = 40745.59) +
  geom_text(aes(x=.87, y=85000, label="High admits, high cost"), color="blue") +
  geom_text(aes(x=.3, y=15000, label="Fewer admits, below average cost"), color="blue") + 
  labs(
    x="Admit rate", 
    y="Average annual cost", 
    title="Exclusive doesn't mean expensive", 
    subtitle="According to the federal Department of Education, there are seven schools who admit fewer than 20 percent of applicants but are among the most affordable.", 
    caption="Source: US Department of Education | By Your Name"
    )
```

```{r bubble4-check}
grade_this_code()
```

This is a good start. It's got some warts -- the grey background, a lack of any real typography, but we're getting to that. Note how I also made the dots for the cheap and exclusive schools red, the bulk of them grey and the lines blue. We're talking about color in the next tutorial.

## The Recap

Throughout this lesson, you've learned how to enhance your data visualizations with key text elements. You've practiced writing headlines that grab attention and convey the main story, crafting subtitles that provide context, and adding clear labels to guide readers through your chart. You've also explored how to use annotations to highlight key insights and divide your chart into meaningful sections. Remember, effective data journalism isn't just about the numbers — it's about using text strategically to tell a compelling story. As you continue to create visualizations, always consider how you can use headlines, labels, and annotations to make your charts more accessible and impactful for your audience.

## Terms to Know

- Headline (Title): The main title of a chart that should convey the key message or story of the visualization.
- Chatter (Subtitle): A brief explanatory text below the headline that provides context or additional information about the chart.
- Annotation: Text or graphical elements added to a chart to highlight or explain specific data points or trends.
- Caption: Text below a chart that provides information about the data source and creator.
- `labs()`: A ggplot2 function used to add various text elements to a chart, including title, subtitle, and axis labels.
- `geom_text()`: A ggplot2 function used to add text directly to a plot at specified coordinates.
- `geom_text_repel()`: A function from the ggrepel package that adds text labels to a plot while avoiding overlaps.
- `geom_hline()`: A ggplot2 function used to add a horizontal line to a plot, often used for showing averages or thresholds.
- `geom_vline()`: A ggplot2 function used to add a vertical line to a plot, often used for showing averages or thresholds.
